---
title: Workflows for association rule mining with staple food crop indicator species
author: M.G. Walsh, R. Manners and M. Broadley
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 1
    fig_caption: true
    css: style.css
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# Introduction

Staples food crops make up the dominant part of a population’s diet. The food staples are eaten regularly supply a major proportion of a person’s energy needs. There are >50,000 edible plants in the world. However, just 15 of these provide ~90% of the world’s *"food energy intake"* (FEI). Rice, maize, and wheat alone make up ~60% of the world's FEI. Other, more regional/local, staples include: millet and sorghum, pulses such as beans, lentils and chickpeas, roots and tubers such as potatoes, cassava, yams, and taro, and animal foods such as meat, fish, and dairy products (see [FAO](http://www.fao.org/3/AC832E/ac832e06.htm)). In turn the specific combinations of staple foods that are consumed by people are thought to have specific population-level nutritional, health, well-being as well as environmental outcomes that are not well described ... and yet ["You are what you eat"](https://www.phrases.org.uk/meanings/you-are-what-you-eat.html).

Association rule mining is frequently used answer exploratory questions like: If someone walks into a supermarket and buys charcoal, what is the probability that that person will also buy steak (or chicken, Bar-B-Que sauce, beer, etc)? This is also why it is frequently referred to as [market basket analysis](https://www.sciencedirect.com/topics/computer-science/market-basket-analysis). East African smallholder food production systems are analogous to a supermarket situation in that small cropland areas (or fields) are generally [intercropped](https://en.wikipedia.org/wiki/Intercropping) with several staple foods, rather than [monocropped](https://en.wikipedia.org/wiki/Monocropping). The small cropland area in East Africa is analogous to a supermarket. Even quite small e.g., 0.25 ha, cropland areas will commonly be planted with multiple staple food crops during any given season. Which specific crops are grown together in a given place or environment, and how these combinations may be evolving due to external factors remains uncertain and unmonitored.

*The main objective of this notebook is to provide starter code for identification of staple food production systems in Rwanda and Tanzania. The data include georeferenced observations of the occurrence of: cattle, sheep, goats, poultry, maize, wheat, sorghum,	rice,	bean,	cowpea,	pigeon pea,	soybean, groundnut, Irish potato,	sweet potato,	cassava, yam, banana, sugarcane, and sunflower. The markdown notebook presented here is maintained on [Github](https://github.com/mgwalsh/RwaSIS/blob/master/Cropping%20_sytems.Rmd), and you can fork and alter it from there for your reference, and as you see fit.*

We will be using [association rule mining](https://en.wikipedia.org/wiki/Association_rule_learning) workflows, based on the concept of strong rules. *"Strong rules"* were introduced by [Agrawal et al., (1993)](https://dl.acm.org/doi/10.1145/170036.170072) in the context of [affinity analysis](https://en.wikipedia.org/wiki/Affinity_analysis). We'll explore these in this notebook to identify cropping combinations (systems) based on staple crop indicator species that can subsequently be used for practical mapping and monitoring applications of staple food crop distributions.

# General data setup

The data we will be using were generated by the [TanSIS]() and [RwaSIS]() projects in Tanzania and Rwanda. They currently consist of >16k georeferenced, time stamped, and photo tagged, field survey observations of staple crops that were observed in 0.25 ha circular plots (~28.2 m radius), in proportion to their national occurrence, between 2015-2021. The spatially representative cropland sampling frames that are used in both currently still ongoing national survey data collections are described in [Walsh, Rutebuka and Manners (2021)](https://github.com/mgwalsh/RwaSIS/blob/master/RwaSIS_cropland_sample.Rmd). A field data collection framework (CropScout) has also been posted on [KoboToolbox](https://kobo.humanitarianresponse.info/#/forms/a3jwzkBTtDvdaZJgR6YjAe/), which you can either use directly or modify for your own use with Open Data Kit ([ODK](https://opendatakit.org/)).

To actually run this notebook, you will need to load the R-packages that are indicated in the chunk directly below to download and assemble the data. You can find all of the documentation of these packages on [CRAN](https://cran.r-project.org/). Note that additional package installs will be needed to run the later sections of this notebook. We shall deal with those as they arise.

```{r}
# Package names
packages <- c("osfr", "tidyverse", "leaflet")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

The survey (label) data downloads can be obtained with this next chunk. It also generates an overview map of where the observations were collected in Tanzania and Rwanda (not shown, but in the markdown document) ... you can click and zoom into the individual locations, which have been surveyed thus far.

```{r, results = 'hide'}
# Create a data folder in  your current working directory
dir.create("Cropping_systems", showWarnings=F)
setwd("./Cropping_systems")
dir.create("Results", showWarnings=F)

# Download crop survey data: https://osf.io/2unc3/
osf_retrieve_file("2unc3") %>% osf_download(conflicts = "overwrite")
staple <- read.table("TZ_RW_staples.csv", header = T, sep = ",")
staple$sdiv <- rowSums(staple[, 6:25]) ## count of number of staples at each survey location
```

```{r, echo = FALSE}
# CropScout sample locations
w <- leaflet() %>%
  setView(lng = mean(staple$lon), lat = mean(staple$lat), zoom = 6) %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  addCircleMarkers(staple$lon, staple$lat, clusterOptions = markerClusterOptions())
w ## plot widget 
```

To prepare the data that are used here for the association rule mining workflow we will need to do a bit wrangling initially to select and recode the variables of interest with the following. The [`arules`](https://www.rdocumentation.org/packages/arules/versions/1.6-8) package, which we will be using further below only processes factor and logical variables. Note that this chunk is a bit ugly, but it works, and we could always convert it into a proper function as more data with different staple crops from different places become available.

```{r}
# Initially select the variables (columns) of interest
staple_ar <- staple[ , c(3:4,1,5:25,27)]

# This does the recoding bits
staple_ar$survey <- as.factor(staple_ar$survey)
staple_ar$build <- as.logical(staple_ar$build)
staple_ar$catt <- as.logical(staple_ar$catt)
staple_ar$shee <- as.logical(staple_ar$shee)
staple_ar$goat <- as.logical(staple_ar$goat)
staple_ar$poul <- as.logical(staple_ar$poul)
staple_ar$maiz <- as.logical(staple_ar$maiz)
staple_ar$whea <- as.logical(staple_ar$whea)
staple_ar$sorg <- as.logical(staple_ar$sorg)
staple_ar$rice <- as.logical(staple_ar$rice)
staple_ar$bean <- as.logical(staple_ar$bean)
staple_ar$cpea <- as.logical(staple_ar$cpea)
staple_ar$pige <- as.logical(staple_ar$pige)
staple_ar$soyb <- as.logical(staple_ar$soyb)
staple_ar$gnut <- as.logical(staple_ar$gnut)
staple_ar$pota <- as.logical(staple_ar$pota)
staple_ar$spot <- as.logical(staple_ar$spot)
staple_ar$cass <- as.logical(staple_ar$cass)
staple_ar$yam <- as.logical(staple_ar$yam)
staple_ar$bana <- as.logical(staple_ar$bana)
staple_ar$cane <- as.logical(staple_ar$cane)
staple_ar$sunf <- as.logical(staple_ar$sunf)

# Provides a glimpse of the resulting dataframe
glimpse(staple_ar)
```

# Association rule discovery

Rule support and confidence are two measures of rule interestingness. They respectively reflect the usefulness and certainty of discovered rules. A support of 2% for Rule (6.1) means that 2% of all the transactions under analysis show that computer and antivirus software are purchased together. A confidence of 60% means that 60% of the customers who purchased a computer also bought the software. Typically, association rules are considered interesting if they satisfy both a minimum support threshold and a minimum confidence threshold. These thresholds can be a set by users or subject matter specialists.


# Association rule mining


# Main takeaways

The main takeaways of this notebook are the following:

* 

Any questions, comments or corrections related to this notebook are most welcome via [AFSIS](mailto:mgw.africasoils.info).

# Aside note: Links to staple food recipes

There are some delicious recipes out there for using the staple crops indicated this notebook for meals (some of these are also quite nutritious). Interestingly cooking advisories such as [Supercook](https://www.supercook.com/#/recipes) and [Allrecipes](https://www.allrecipes.com/search/results/?search=) are also using new recommendation algorithms to help provide meal options based on what ingredients you have available, or may want to have available, in your fridge and dry-goods store. If you are a smallholder in East Africa, those ingredients will largely depend on what you are growing on your land and whatever your breakfast, lunch or dinner preferences are. Maybe we (data scientists) should start to work back from what smallholders are currently growing and the attached preferences?


