---
title: Workflows for mapping cropping systems using staple food indicator species
author: M.G Walsh, R. Manners, J. Meliyo, and J. Rutembuka
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The term ["Agroecology"](https://www.agroecologyfund.org/what-is-agroecology) refers to farming that “centers on food production that makes the best use of nature’s goods and services while not damaging those resources.” Agroecological practices are closely linked to knowledge about the crop (and livestock) combinations, crop sequences and management techniques i.e., ["Cropping systems"](https://www.sciencedirect.com/topics/agricultural-and-biological-sciences/cropping-system) that are applied on a given field, farm or region over a period of years. The prevailing cropping systems in East Africa often stand in stark contrast to industrial agricultural practices that rely on reduced labor with mechanization, consolidation of farms and fields, and increased external agricultural inputs such as fossil fuels, electricity, irrigation, GMOs, improved / patented seed, chemical fertilizers, pesticides and herbicides. The increased reliance on external inputs in turn usually comes at economic, health / nutritional, environmental and/or [Ecosystem service](https://en.wikipedia.org/wiki/Ecosystem_service) costs to smallholder farmers and rural communities. It is not possible to gauge or manage the ensuing agroecological interactions and complexities without knowing where specific cropping systems and management practices are located and how they may be evolving.

*The objective of this notebook is to provide starter code in [R](https://www.r-project.org/) for classifying and delineating of [Staple food cropping systems](https://en.wikipedia.org/wiki/Staple_food) in Rwanda and Tanzania. The data include georeferenced observations of: maize, millet, wheat, sorghum, rice, bean, chickpea, pigeon pea, soybean, Irish potatoes, sweet potatoes, cassava and banana, in addition to livestock as well as the combined presence/absence of numerous other, more localized food crops. The markdown notebook presented here is maintained on [Github](), and you can fork and alter it from there for your reference and as you see fit.*

# General data setup

The data we will be using were generated by the [TanSIS]() and [RwaSIS]() projects located in Tanzania and Rwanda respectively. They consist of georeferenced field survey observations of staple crops observed in 1 ha circular plots (~56.42 m radius), and in proportion to their national occurrence. The nationally representative cropland sampling frames that are used in the survey data collections are described in [Walsh et al., (2021)](). A field data collection framework (CropScout) has been posted on [KoboToolbox](https://kobo.humanitarianresponse.info/#/forms/a3jwzkBTtDvdaZJgR6YjAe/), which you can either use directly or modify for your own use in Open Data Kit [ODK](https://opendatakit.org/).

## Packages and data download

To actually run this notebook, you will need to load the packages indicated in the chunk directly below to download and assemble the data. The chunk (not shown but in the markdown document) also generates an overview map of where the observations were collected in Tanzania and Rwanda ... you can click and zoom into the individual locations.

```{r, results = 'hide'}
# Package names
packages <- c("osfr", "dplyr")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

