---
title: Meta-analyses of agricultual lime application effects on crop yields with field trial data
author: M.G. Walsh, R. Manners, J.V. Silva and S. Aston
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    css: style.css
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# About this notebook

Soil acidity affects soil nutrient availability and microbial activity and is thought to reduce crop growth and yield. For example, in Rwanda virtually all of the currently surveyed cropland soils are acid (pH < 7 in water) and ~64% are strongly acid (with pH < 5.5); also see the associated notebooks about Rwanda's [Cropland soil pH](https://github.com/mgwalsh/Soils/blob/master/PSM_guide.Rmd) and its [Crop distribution](https://github.com/mgwalsh/RwaSIS/blob/master/RW_priority_crops.Rmd). Moreover, commonly used nitrogen fertilizers that supply ammonium or that react in the soil to produce ammonium (e.g., ammonium nitrate, urea, manure and compost) will tend to decrease soil pH over time. The application of [Agricultural lime (aglime)](https://en.wikipedia.org/wiki/Agricultural_lime) can be used to mitigate the effects of soil acidity on soil nutrient availability, reduced microbial activity, and the occurrence of aluminium and/or manganese toxicities (e.g., see the extension advice from [PennState Extension](https://extension.psu.edu/soil-acidity-and-aglime)). The rocks and minerals from which agricultural lime is derived, primarily [Limestone](https://en.wikipedia.org/wiki/Limestone), but also [Chalk](https://en.wikipedia.org/wiki/Chalk), is primarily composed of calcium carbonate (CaCO~3~) and/or magnesium carbonate (MgCO~3~). Limestone is typically pulverized and/or chemically altered to produce different aglime products of varying efficacy. Similar to the application of fertilizers, the use of aglime also generates significant greenhouse gas emissions in its production, to-farm-gate transport and through its field application, that should be minimized on that basis.

*The main objective of this notebook is to illustrate the use of standard meta-analysis techniques for assessing the effects of aglime applications on maize and bean yields in Rwanda. Rwanda was selected for this illustration because of the high prevalence of acid soils within its croplands, as well as the availability of a relatively large, recent, multi-year, field trial dataset that is suitable for this purpose. The notebook itself is maintained on [Github](https://github.com/mgwalsh/RwaSIS/blob/master/Meta_analysis_field_trials.Rmd), and you can fork and modify it from there as you see fit.*

A [Meta-analysis (MA)](https://en.wikipedia.org/wiki/Meta-analysis) is an analytical approach that uses statistical techniques to combine the results from different studies to obtain quantitative estimates of the overall effects of particular interventions or variables on defined outcomes. Generally, the aim is to derive a pooled estimate of a common outcome, based on the relevant sources of variation across studies. The scope of MAs is defined by the type of population, the types of interventions (and their controls or comparators), and the types of outcomes that are of interest. The acronym PICO (population, interventions, comparators and outcomes) helps to serve as a reminder of these (e.g., see the [Cochrane Handbook](https://training.cochrane.org/handbook)). We will follow the Cochrane guidance here and will go through some of the relevant analysis steps including exploratory data analysis ([EDA](https://en.wikipedia.org/wiki/Exploratory_data_analysis)), [Outlier detection](https://en.wikipedia.org/wiki/Anomaly_detection), [Data modeling](https://en.wikipedia.org/wiki/Data_modeling) and interpretation.

# General data set up

To actually run this notebook, you will need to load the packages indicated in the chunk directly below to download, assemble and analyze the Rwanda lime trial data.

```{r, results = 'hide'}
# Package names
packages <- c("osfr", "dplyr", "table1", "solitude", "MASS", "stargazer", "arm")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

The next chunk downloads eligibility pre-screened trial data from the [RwaSIS OSF repository](https://doi.org/10.17605/OSF.IO/Y9ZUT). The raw data are also openly available from [1AF](mailto:step.aston@oneacrefund.org ) on request, if you would like to trace our initial data screening and wrangling processes.

```{r, results = 'hide'}
# Create a data folder in your current working directory
dir.create("RW_lime", showWarnings=F)
setwd("./RW_lime")
dir.create("Results")

# Trial data download from OSF at: https://osf.io/ngmjy/
osf_retrieve_file("ngmjy") %>% osf_download(conflicts = "overwrite")
lime <- read.table("RW_1AF_lime.csv", header=T, sep=",")
```

# Population, interventions, comparators and outcome(s) (PICO)

The data we will be using in this notebook were generated by the [One Acre Fund (1AF)](https://oneacrefund.org/). Two separate field studies were conducted by 1AF as part of their soil acidity management intervention testing and monitoring activities in Rwanda during 2016-2017 and 2019-2020. They consist of 2,166 individual (multi-location, on-farm) trials with n = 10,526 crop yield (kg/ha) measurments × `crop` × `lime`, to assess the effects of aglime applications on crop (maize & bean) yields against the relevant control yields i.e., those without lime application. The 2 studies were conducted in 20 Level-4 administrative units (or so-called "Cells" in Rwanda), two of which overlap temporally by `study`. Crop yield measurements were generally repeated at least over 2 cropping seasons in each study, defined by the relevant `year` × `cell` × `crop` × `lime` × `fert` treatment combinations. Note that this also defines the main **PICO** data that are analyzed here.

## Table 1

It is common practice in meta-analysis that the first table of e.g., a [Cochrane review article](https://www.cochranelibrary.com/), referred to as “Table 1”, presents the descriptive statistics of the outcome characteristics of the study population that is stratified by the main individual treatment / comparator variables. The [table1](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html) package in R is useful and easy in this context. We use it here to illustrate both the marginal treatment and random effects (`year` × `cell`) on maize and bean crop yields. The following chunk generates the table ... you can modify it as you see fit.

```{r}

```

## PICO yield outlier screening

# Data modeling

# Takeaways

The main takeaways of this notebook are the following:

* 

Any questions, comments or corrections related to this notebook are most welcome via [AFSIS](mailto:mgw.africasoils.info).

# Recommended trial protocol adjustments for 1AF

* Put all raw data on KoboToolbox. 

* Georeference time stamp and photograph individual field trials.

* Record intercrops.

* Repeat soil acidity (pH & exchangeable acidity) measurements over time.

* Systematically document crop varieties.

* Specify lime provenance data [Lime sources]()

* Generate efficient response trial designs. 

* Put all of the analysis code on Github and/or OSF.
