---
title: Meta-analyses of agricultual lime application effects on crop yields from field trial data
author: M.G. Walsh, R. Manners, J.V. Silva, S. Gameda and S. Aston
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    css: style.css
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction

Soil acidity affects soil nutrient availability and microbial activity and is generally thought to reduce crop growth and yield. For example, in Rwanda virtually all of the currently surveyed cropland soils are acid (pH < 7 in water) and ~64% are strongly acidic (with pH < 5.5). Also see the associated notebooks about Rwanda's [Cropland soil pH](https://github.com/mgwalsh/Soils/blob/master/PSM_guide.Rmd) and its [maize & bean distributions](https://github.com/mgwalsh/RwaSIS/blob/master/RW_priority_crops.Rmd). Moreover, commonly used nitrogen fertilizers that supply ammonium directly or that react in the soil to produce ammonium, such as ammonium nitrate, urea, manure and compost, will decrease soil pH over time (also see the extension advice provided by [PennState Extension](https://extension.psu.edu/soil-acidity-and-aglime)). 

The application of [Agricultural lime (aglime)](https://en.wikipedia.org/wiki/Agricultural_lime) can be used to mitigate the effects of soil acidity on nutrient availability, reduced microbial activity, and the occurrence of [Aluminium](https://www.sciencedirect.com/science/article/pii/B9780128031582000011) and/or [Manganese](https://www.tandfonline.com/doi/abs/10.1080/01904169809365409) toxicities. The rocks and minerals from which agricultural lime is derived, primarily [Limestone](https://en.wikipedia.org/wiki/Limestone), but also [Chalk](https://en.wikipedia.org/wiki/Chalk), is primarily composed of calcium (CaO) and/or magnesium (MgO) oxides, which are readily measured by e.g., [X-ray fluorescence](https://en.wikipedia.org/wiki/X-ray_fluorescence). Limestone is typically pulverized and/or chemically altered to produce different aglime products of varying efficacy in increasing soil pH. Similar to chemical fertilizers, the use of aglime also generates significant greenhouse gas emissions in its production, to-farm-gate transport and through its field application, which should be minimized on that basis.

*The main objective of this notebook is to illustrate the use of standard meta-analysis techniques for assessing the effects of aglime applications on maize and bean yields in Rwanda. Rwanda was selected for this illustration because of the high prevalence of acid soils within its croplands, as well as the availability of a relatively large, recent, multi-year, field trial dataset that is suitable for this purpose. The notebook itself is maintained on [Github](https://github.com/mgwalsh/RwaSIS/blob/master/Meta_analysis_field_trials.Rmd), and you can fork and modify it from there as you see fit.*

A [Meta-analysis (MA)](https://en.wikipedia.org/wiki/Meta-analysis) is an analytical approach that uses statistical techniques to combine the results from different studies to obtain quantitative estimates of the overall effects of particular interventions or variables on defined outcomes. Generally, the aim is to derive a pooled estimate of a common outcome, based on the relevant sources of variation across studies. The scope of MAs is defined by the type of population, the types of interventions (and their controls or comparators), and the types of outcomes that are of interest. The acronym PICO (population, interventions, comparators and outcomes) helps to serve as a reminder of these (e.g., see the [Cochrane Handbook](https://training.cochrane.org/handbook)). We will follow the Cochrane guidance here and will go through some of the relevant analysis steps including exploratory data analysis ([EDA](https://en.wikipedia.org/wiki/Exploratory_data_analysis)), [outlier detection](https://en.wikipedia.org/wiki/Anomaly_detection), [data modeling](https://en.wikipedia.org/wiki/Data_modeling) and interpretation.

# Data set up

To actually run this notebook, you will need to load the packages indicated in the chunk directly below to download, assemble and analyze the Rwanda lime trial data.

```{r, results = 'hide'}
# Package names
packages <- c("osfr", "dplyr", "MASS", "stargazer", "arm")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

The next chunk downloads eligibility pre-screened trial data from the [RwaSIS OSF repository](https://doi.org/10.17605/OSF.IO/Y9ZUT). The raw data are also openly available from [1AF](mailto:step.aston@oneacrefund.org) on email request, if you would like to trace our initial data screening and wrangling efforts.

```{r, results = 'hide'}
# Create a data folder in your current working directory
dir.create("RW_lime", showWarnings=F)
setwd("./RW_lime")
dir.create("Results")

# Trial data download from OSF at: https://osf.io/ngmjy/
osf_retrieve_file("ngmjy") %>% osf_download(conflicts = "overwrite")
lime <- read.table("RW_1AF_lime.csv", header=T, sep=",")
```

The below is the structure of the resulting dataframe that we shall be using. The only variable, which has not been properly cleaned yet in these data is the crop variety indicator (`crop_var`). We will not be using that in this notebook. However, we have included it for reference and any potential updates in future versions of this notebook (e.g., see [Jaleta et al., 2020](https://doi.org/10.1371/journal.pone.0235484) as to why this might be important).

```{r, echo = FALSE}
str(lime)
```

# Population, interventions, comparators and outcome (PICO)

The data we will be using in this notebook were generated by the [One Acre Fund (1AF)](https://oneacrefund.org/). Two separate field studies were conducted by 1AF as part of their soil acidity management intervention testing and monitoring activities in Rwanda during 2016-2017 and 2019-2020. They consist of 2,166 individual (multi-location, on-farm) trials with n = 10,464 crop yield (kg ha^-1^) measurements × `crop` × `lime` × `fert` treatments, to assess the effects of aglime applications on crop yields against the relevant control yields i.e., those without lime application. The 2 studies were conducted in 20 Level-4 administrative units (or so-called "Cells" in Rwanda), two of which overlap temporally by `study`. Crop yield measurements were generally repeated over at least 2 cropping seasons in each study, defined by the relevant `year` × `cell` × `crop` × `lime` × `fert` treatment combinations. Note that this combination also defines the main **PICO** that are analyzed here.

## PICO Tables

It is common practice in meta-analysis that the first table of e.g., a [Cochrane review article](https://www.cochranelibrary.com/), often referred to as “Table 1”, summarizes the descriptive statistics of the characteristics of the study populations that are stratified by the main individual treatment / comparator variables. The [`aggregate`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/aggregate) function in the R [`stats`](https://www.rdocumentation.org/packages/stats/versions/3.6.2) package is convenient in this context. We use it here to summarize both the fixed treatment (`crop` × `lime` × `fert`) and the associated random effects (`trial_age` × `cell`) that would go into the model specification of a [Production function](https://en.wikipedia.org/wiki/Production_function). The following chunks generate the associated example crop yield quantile tables.

```{r}
# main fixed effects on yields (kg / ha)
fix <- aggregate(lime$yield, list(lime$lime, lime$fert, lime$crop), FUN = quantile, probs = c(0.05, 0.5, 0.95))
colnames(fix) <- c("lime", "fert", "crop", "Q")
fix
```

```{r}
# random effects on yields (kg / ha)
ran <- aggregate(lime$yield, list(lime$trial_age, lime$cell), FUN = quantile, probs = c(0.05, 0.5, 0.95))
colnames(ran) <- c("trial_age", "cell", "Q")
ran
```

## PICO outlier screening

The PICO tables (above) show some quite high crop yield outliers that should be screened before proceeding to any data modeling and/or machine learning steps. There are also zero yields, which may be due to censoring, lack of follow-up and/or local crop failures. However, the causes of these are not discernible from the data. There are also a systematic trial drop-outs in the crop yield data follow-ups with `trial_age` in both studies based on ... 

```{r}
tri <- aggregate(lime$yield, list(lime$trial_age, lime$study), FUN = length)
colnames(tri) <- c("trial_age", "study", "n")
tri
```

However, because of the lack georeferencing of the individual trials it is not possible to tell which trials dropped-out during the second year in each study. Indeed, new trials may also have come into the studies during the follow-up period, that were not recorded as such ... but we have no way of telling. This might also have affected the somewhat counter-intuitive crop yield differences that were recorded by time-period in the 2 studies.

```{r}
tri <- aggregate(lime$yield, list(lime$year, lime$crop), FUN = quantile, probs = c(0.05, 0.5, 0.95))
colnames(tri) <- c("year", "crop", "Q")
tri
```


# Data modeling

# Takeaways

The main takeaways of this notebook are the following:

* 

Any questions, comments or corrections related to this notebook are most welcome via [AFSIS](mailto:mgw.africasoils.info).

# Recommended trial protocol adjustments

* Put all raw survey and experiment data on KoBoToolbox. Just go to [KoBoToolbox](https://www.kobotoolbox.org/), set it up in Excel (or anything else) and deploy it for your surveys and experiments using [ODK](https://opendatakit.org/) ... both are free and easy to use! Then georeference, time stamp, and photograph all individual field trials and survey locations. 

* East Africa is certainly not Nebraska or Brazil. Intercrops occur virtually everywhere in Rwanda's smallholder production systems and elsewhere in East Africa. Our recommendation is to start designing your agricultural input trials in ways that actually reflect those facts rather than on the basis of monocropping practices.

* Repeat soil acidity (pH & exchangeable acidity) measurements over time. You don't really need a spectrometer to do that, though it might help. [pH probes](https://www.amazon.com/ph-probe/s?k=ph+probe) sell for < 50 U$ on Amazon.

* Systematically document the crop varieties that are used in trials. This would help enormously in assessing genotype × environment interactions ([GE](https://en.wikipedia.org/wiki/Gene%E2%80%93environment_interaction)) that plant breeders need to assess, recommend, and develop acid tolerant crop varieties in different places. 

* Specify lime provenance data (see e.g., [Rwanda aglime sources]()). Aglime sources in Rwanda and elsewhere are quite variable in terms of their calcium cabonate equivalents (CCE), Ca/Mg ratios and estimated extractable limestone source reserves.

* Generate more efficient agronomic response trial designs. 

* Put all of the analysis code on Github and/or OSF! Make the data and the code truly [FAIR](https://en.wikipedia.org/wiki/FAIR_data) so that others can attempt to help you more easily!
