---
title: Crop distribution survey meta-analyses and small area estimation 
author: M.G. Walsh, R. Manners, ...
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 1
    fig_caption: true
    css: style.css
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# Introduction

The main intent of this notebook is to illustrate starter code for predictive crop type and cropping systems distribution mapping and the statistical small area estimates [SAE](https://www.census.gov/srd/csrm/SmallArea.html) based on that, which define the crop and cropping systems types in a given country or any other ROI. We use both Rwanda's most recent and some of its legacy cropland survey (label) data here and the associated gridded raster (features) to illustrate the general approach and the main data analysis steps. Rwanda, being a small country, is convenient for this illustration because the script will run fast so as to not test your patience in running it. The notebook is publicly maintained on [Github](https://github.com/mgwalsh/RwaSIS/blob/master/RW_priority_crops.Rmd), and you can fork and modify it from there as you see fit. You may also want to take a look at the [GeoSurvey cropland area notebook](https://osf.io/shkxp/), which describes some of the spatial prediction methods and the associated SAE procedures in more detail.

# General data setup

To actually run this notebook, you will need to load the packages indicated in the chunk directly below. This allows you to assemble Rwanda-wide observations, link those to the spatial data and then model them using machine learning. 

```{r}
# Package names
packages <- c("downloader", "jsonlite", "rgdal", "sp", "raster", "leaflet", "DT", "htmlwidgets", "devtools", "caret", "caretEnsemble", "mgcv", "MASS", "randomForest", "xgboost", "nnet", "dplyr", "doParallel", "dismo", "arm")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
    install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

## Data downloads

The following chunk downloads the data that we have assembled, which are needed for running this particular example. Specifically, it assembles georeferenced field observations of priority crops that have been requested by [RAB](http://rab.gov.rw/index.php?id=180), Maize, Wheat, Rice, Potato, Cassava and Bean, for their planned management response trials and links these to remote sensing images and GIS data represented by the `grids` feature stack. We focus on predicting the distribution of Maize here, but the approach is equally applicable to any given  [georeferenced](https://en.wikipedia.org/wiki/Georeferencing) crop type under consideration. Note that this chunk is Mac or Linux specific and so the directory structures should be changed to run on Windows machines.

```{r}
# Data downloads -----------------------------------------------------------
# Create a data folder in  your current working directory
dir.create("Priority_crops", showWarnings=F)
setwd("./Priority_crops")
dir.create("Results", showWarnings=F)

# Download crop data
download("", mode = "wb")
samp <- read.table("Samples.csv", header = T, sep = ",")

# Download raster stacks
download("https://osf.io/hp6v7?raw=1", "RW_250m_2020.zip", mode = "wb")
unzip("RW_250m_2020.zip", overwrite = T)
download("https://osf.io/u73pd?raw=1", "RW_GS_preds.zip", mode = "wb")
unzip("RW_GS_preds.zip", overwrite = T)
glist <- list.files(pattern="tif", full.names = T)
grids <- stack(glist)

# Download cropland mask
download("https://osf.io/bmysp/", "RW_CP_mask.zip", mode = "wb")
exdir <- "./Mask" 
unzip("RW_CP_mask.zip", exdir = exdir, overwrite = T)
```

## Observational data

RAB's 6 priority crop (presence/absence) observations for Rwanda were assembled from 4 surveys:

1. The currently ongoing RwaSIS survey designated as `base` (see and download the raw data at: [Kobo](https://kobo.humanitarianresponse.info/#/forms/a3jwzkBTtDvdaZJgR6YjAe/summary)). There are currently 1,126 observations in this survey in which at least one of the priority crops is present. Note that this number will change during 2021/22 as the RwaSIS survey progresses and we will update the numbers accordingly.

2. The [CIALCA Farm Heterogeneity Survey of Rwanda](https://cialca.shinyapps.io/cialca_base_2files/) survey desinated as `s1`. There are 2,483 georeferenced observations in this survey in which at least one of the priority crops is present. Note that the spatial coordinates for this survey have been slightly offset to protect the privacy of farmers.

3. The [CNLS]() soil survey of Rwanda's croplands, designated as `s2`, which is openly accessible at: [OSF](). There are 602 observations in this survey in which at least one of the priority crops is present. These were extracted from the crop type descriptions from the soil survey meta-data.

4. The [RAB]() soil survey of lowland rice areas of Rwanda, designated as `s3`. There are 916 observations in this survey for sites where only rice was present, by design. Note that this is an [oversample](https://en.wikipedia.org/wiki/Oversampling_and_undersampling_in_data_analysis) of the rice growing area of Rwanda and has to be treated as such in some of the analyses below. 

## Gridded data

The pre-processed Rwanda grid data (in the `grids` raster stack) were derived and reprojected from their primary open sources. You can also download the entire stack directly at [RwaSIS grids](https://osf.io/hp6v7/). The short descriptions of the included rasters, and their sources are provided in the table immediately below.

\
```{r, echo=FALSE, results='asis'}
download("https://osf.io/ag6fk?raw=1", ".//GS variables.csv", mode = "wb")
vars <- read.table("./Priority_crops/GS variables.csv", header = T, sep = ",")
datatable(vars)
```
\
These Rwanda-wide (actually Africa-wide) features will change over time and I will update them if and when needed. Also note that these are grouped by factor variables that designate the occurrence of a given crop as a [Jenny-type](https://soilandhealth.org/wp-content/uploads/01aglibrary/010159.Jenny.pdf) function *f*(C) ~ (*a, c, o, r, s*) where:

* a - anthropic variables
* c - climatic variables
* o - organismal and successional variables
* r - relief / topographical variables
* s - parent material and soil related variables

The main notion here is that the occurrence and distribution of croplands and crop types must always be associated with the distribution of humans and their built infrastructure, but also constrained or facilitated by changes in typically much slower environmental factors such as climate, [ecological succession](https://en.wikipedia.org/wiki/Ecological_succession), topography, parent materials and soils. Note that all of these change and interact over space-time and should be, but are currently not adequately monitored across Africa.
