---
title: RwaSIS cropland sampling frame
author: M.G. Walsh, J. Rutebuka and R. Manners
date: "`r format(Sys.time(), '%d, %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 1
    css: style.css
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# Introduction

This notebook describes setting up a representative and completely reproducible [sampling frame](https://en.wikipedia.org/wiki/Sampling_frame) that provides spatially/geographically balanced coverage across the croplands of Rwanda. Croplands are the primary [Region of Interest (ROI)](https://en.wikipedia.org/wiki/Region_of_interest) and the target for various land management interventions of the [RwaSIS project](https://minagri.prod.risa.rw/updates/news-details/rwanda-launches-soil-information-service-project-to-improve-agriculture-productivity). Based on a recent high-resolution remote-sensing [GeoSurvey (2019)](https://osf.io/w5e6c/), croplands are currently estimated to occupy ~68% of Rwanda's overall land area (~2.37 Mha). Selecting an appropriate sampling frame for ground observations and measurements and/or experiments is a critical planning step for the RwaSIS project, because it determines both the main and recurrent costs of any adaptive mapping or monitoring program as well as the resulting precision and accuracy of inferences and predictions that are made about cropland condition and productivity over time. A geographically consistent sampling frame for field surveys and experiments should therefore always be defined in terms of the immediate needs and constraints of the measurement, mapping and monitoring tasks at hand, but within a long-term planning context.

# RwaSIS cropland sampling frame R-script

These are the R-packages that you will need to run to reproduce or adapt the spatially balanced sampling script that we are proposing. The notebook itself is maintained on Github [here](https://github.com/mgwalsh/RwaSIS/blob/master/RwaSIS_cropland_sample.Rmd). You can fork it from there and modify it as you see fit.

```{r}
# Package names
packages <- c("rgdal", "raster", "sp", "BalancedSampling", "DT", "leaflet", "htmlwidgets")

# Install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

The following chunk downloads the needed and publicly available geotiffs and shape file that define the RwaSIS project's main ROI constraints (i.e, cropland && distance to buildings < 0.5 km).

```{r}
# Data setup --------------------------------------------------------------
# create a data folder in your current working directory
dir.create("RW_MS_sample", showWarnings=F)
setwd("./RW_MS_sample")

# download GADM-L5 shapefile (courtesy: http://www.gadm.org)
download.file("https://www.dropbox.com/s/fhusrzswk599crn/RWA_level5.zip?raw=1", "RWA_level5.zip")
unzip("RWA_level5.zip", overwrite=T)
shape <- shapefile("gadm36_RWA_5.shp")

# download GeoSurvey prediction layers
download.file("https://osf.io/u73pd?raw=1", "RW_GS_preds.zip")
unzip("RW_GS_preds.zip", overwrite=T)
glist <- list.files(pattern="tif", full.names=T)
grids <- stack(glist)
```

The subsequent chunk then then sets-up the layers for drawing a spatially balanced survey location sample from the ROI. We use the `overlay` and `lcube` functions from the `raster` (to create the ROI) and `BalancedSampling` (to sample the ROI) libraries in R [(R Core Team)](https://cran.univ-paris1.fr/web/packages/dplR/vignettes/intro-dplR.pdf) to draw a spatially representative sample. The method implements the cube method of [Deville and Tillé (2004)](https://academic.oup.com/biomet/article-abstract/91/4/893/459329?redirectedFrom=fulltext), which allows sampling based on the relevant inclusion probabilities while aiming for balance and spread with respect to speciﬁed covariates and/or constraints. 

```{r}
# Sample setup ------------------------------------------------------------
# create a ROI image based on cropland mask and distance to nearest buildings
cp <- 1    ## set cropland mask to 1 (present)
bd <- 0.5  ## set maximum distance to the nearest "buildings" (in km) ... for logistical reasons
roi <- overlay(grids, fun=function(x) 
{return(ifelse(x[4] == cp && x[2] <= bd, 1, 0))}) ## extracts the ROI

# extract ROI coordinates
coord <- coordinates(roi)
index <- extract(roi, coord)
index <- as.data.frame(cbind(coord, index))
rmask <- index[which(index$index == 1),]
```

In this case potential survey sites falling within the ROI were selected purely for spatial balance, which entails that the mean coordinates of sample sites are close to the mean coordinates of all points in the sample frame and have adequate spatial spread. This ensures that the observations are spread out rather than clustered with respect to the spatial coordinates, see [Grafström and Schelin (2014)](https://cran.r-project.org/web/packages/BalancedSampling/index.html). This next chunk then draws the spatially balanced sample of the proposed cropland sampling locations.

```{r}
# Spatially balanced sampling ---------------------------------------------
# set sampling parameters
N <- nrow(rmask) ## ROI size (in 250 m pixels)
n <- round(N/16*0.15,0) ## set sample size (number of sampling locations)
p <- rep(n/N,N)  ## inclusion probabilities

# draw geographically balanced sample
set.seed(6405)                      ## sets a randomization seed, note if you change this you will get different results
B <- cbind(p, rmask[,1], rmask[,2]) ## specifies spatial balancing variables
rsamp <- cube(p, B)                 ## samples from the ROI
```

The next chunk generates various output files, which can be used for field survey planning and navigation.

```{r}
# Write files -------------------------------------------------------------
# extract sample coordinates
x <- rmask[rsamp,1]
y <- rmask[rsamp,2]
xy <- data.frame(cbind(x,y))

# Define unique grid ID's (GID)
# Specify GID scale (res.pixel, in m)
res.pixel <- 10000

# Grid ID (GID) definition
xgid <- ceiling(abs(xy$x)/res.pixel)
ygid <- ceiling(abs(xy$y)/res.pixel)
gidx <- ifelse(xy$x<0, paste("W", xgid, sep=""), paste("E", xgid, sep=""))
gidy <- ifelse(xy$y<0, paste("S", ygid, sep=""), paste("N", ygid, sep=""))
GID <- paste(gidx, gidy, sep="")
xy <- cbind(GID, xy)

# attach GADM-L5 and above unit names from shape
coordinates(xy) <- ~x+y
crs(xy) <- "+proj=laea +ellps=WGS84 +lon_0=20 +lat_0=5 +units=m +no_defs"
sloc <- spTransform(xy, CRS(proj4string(shape)))
gadm <- sloc %over% shape
sloc <- as.data.frame(sloc)
samp <- cbind(gadm[ ,c(4,6,8,10,12)], sloc)
colnames(samp) <- c("province","district","sector","cell","village","gid","lon","lat")
write.csv(samp, "RW_cropland_sample.csv", row.names = F)
```

The table below is the main output data frame `samp` of the proposed sampling locations with their lon/lat coordinates, which can be sorted by various administrative units and/or `GID`. The `GID` is sort of like a military style 10x10 km grid ID that will usually contain several sampling locations. Our practice has been to prioritize survey activities in `GID` that contain at least several locations. The data frame is also written as a csv file `RW_cropland_sample.csv` into your working directory, which you can use as waypoint input to a GPS (see e.g: [GPSBabel](https://www.gpsbabel.org)), tablet or smart phone for in-the-field navigation.

\
```{r, echo=FALSE}
datatable(samp)
```
\
You can also generate a zoomable map of the proposed survey locations with the `leaflet` package. This shows the proposed distribution of the survey sites, which will help with navigation and operational planning on the ground. It is also saved in your working directory and can be used in presentations and additional planning documents.

```{r}
# Sampling map widget
# render map
w <- leaflet() %>%
  setView(lng = mean(samp$lon), lat = mean(samp$lat), zoom = 8) %>%
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  addCircleMarkers(samp$lon, samp$lat, clusterOptions = markerClusterOptions())
saveWidget(w, "RwaSIS_cropland_sample.html", selfcontained = T) ## save map widget
w ## plot widget 
```

# Follow-up on the ground

Note that follow-up on the ground will be needed to actually install survey locations and/or field trials. This will require negogiations with farmers, government, private sector and civil society representatives at the proposed locations and perhaps also a pre-survey of the prevailing cropping systems and soil condition for such surveys or experiments. We will provide additional guidance about the relevant field procedures and potential experimental setups ... stay tuned.
